---
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
library(sf)
library(ggplot2)
library(ggspatial)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)
```

#### 2. **Load Rwanda shapefile and Gishwati-Mukura reserve shapefile**

If you don’t already have shapefiles:Get **Rwanda** boundaries from
`rnaturalearth`.

-   Use shapefiles for **Gishwati-Mukura** Biosphere (download from a
    reliable GIS source or governmental portal).

```{r}


# Set Rwanda-specific projection (UTM Zone 35S)
rwanda_crs <- 32735  # EPSG:32735 for Rwanda (WGS 84 / UTM zone 35S)

# Get Rwanda map and reproject
rwanda <- ne_countries(scale = "medium", country = "Rwanda", returnclass = "sf") %>%
  st_transform(crs = rwanda_crs)

# Set directory path
shapefile_dir <- "/Users/blaise/Rwanda_National Parks/Code_Species_Modeling/Gishwati_Mukura"

# Read and reproject Gishwati and Mukura shapefiles
gishwati <- st_read(file.path(shapefile_dir, "Gishwati_core.shp")) %>%
  st_transform(crs = rwanda_crs)
mukura <- st_read(file.path(shapefile_dir, "Mukura_core.shp")) %>%
  st_transform(crs = rwanda_crs)
gishwati_buffer <- st_read(file.path(shapefile_dir, "Gishwati_buffer.shp")) %>%
  st_transform(crs = rwanda_crs)
mukura_buffer <- st_read(file.path(shapefile_dir, "Mukura_buffer.shp")) %>%
  st_transform(crs = rwanda_crs)


```

```{r}
library(dplyr)
gishwati <- gishwati %>% mutate(source = "Gishwati_core") %>% select(source, geometry)
mukura <- mukura %>% mutate(source = "Mukura_core") %>% select(source, geometry)
gishwati_buffer <- gishwati_buffer %>% mutate(source = "Gishwati_buffer") %>% select(source, geometry)
mukura_buffer <- mukura_buffer %>% mutate(source = "Mukura_buffer") %>% select(source, geometry)

# Combine all with a source column
gmb_reserve <- rbind(gishwati, mukura, gishwati_buffer, mukura_buffer)

```

```{r}
unique(st_geometry_type(gmb_reserve))




```

```{r}
library(sf)
library(dplyr)

# 1. Set Rwanda-specific projection (WGS 84 / UTM zone 35S)
#rwanda_crs <- 32735  # EPSG code

# 2. Set shapefile directory path
#shapefile_dir <- "/Users/blaise/Rwanda_National Parks/Code_Species_Modeling/Gishwati_Mukura"

# 3. Read all necessary shapefiles
rwanda_boundary <- st_read(file.path(shapefile_dir, "Rwanda_boundary.shp"))
lakes           <- st_read(file.path(shapefile_dir, "Lakes.shp"))
lake_kivu       <- st_read(file.path(shapefile_dir, "Lake_Kivu.shp"))
volcanoes       <- st_read(file.path(shapefile_dir, "Volcanoes.shp"))
akagera         <- st_read(file.path(shapefile_dir, "Akagera.shp"))

# 4. Reproject all to Rwanda UTM CRS (EPSG:32735)
rwanda_boundary <- st_transform(rwanda_boundary, crs = rwanda_crs)
lakes           <- st_transform(lakes,           crs = rwanda_crs)
lake_kivu       <- st_transform(lake_kivu,       crs = rwanda_crs)
volcanoes       <- st_transform(volcanoes,       crs = rwanda_crs)
akagera         <- st_transform(akagera,         crs = rwanda_crs)

# 5. Store all as named layers for easy access
rwanda_layers <- list(
  boundary   = rwanda_boundary,
  lakes      = lakes,
  lake_kivu  = lake_kivu,
  volcanoes  = volcanoes,
  akagera    = akagera
)

# 6. If you want a simple variable called `rwanda` for mapping:
rwanda <- rwanda_boundary

```

```{r}
library(ggplot2)

ggplot() +
  geom_sf(data = rwanda_layers$boundary, fill = "white", color = "black") +
  geom_sf(data = rwanda_layers$lakes, fill = "lightblue", color = NA) +
  geom_sf(data = rwanda_layers$lake_kivu, fill = "lightblue", color = NA) +
  geom_sf(data = rwanda_layers$akagera, fill = "darkgreen", alpha = 0.3) +
  geom_sf(data = rwanda_layers$volcanoes, color = "darkgreen", size = 1) +
  coord_sf(crs = rwanda_crs) +
  theme_minimal() +
  labs(title = "Rwanda Map with Key Features")

```

```{r}
# Read FIA plot data from CSV
FiaPlot_dir <- "/Users/blaise/Rwanda_National Parks/Code_Species_Modeling/code"
fia_csv <- read.csv(file.path(FiaPlot_dir, "groundtruth_data.csv"))

str(fia_csv)

```

Add a negative sign to those columns for only "Gishwati"

```{r}
library(dplyr)

fia_csv <- fia_csv %>%
  mutate(across(
    .cols = c(Corner_1.Latitude,  Corner_2.Latitude, Corner_3.Latitude, Corner_4.Latitude),
    .fns = ~ ifelse(Site == "Gishwati", -abs(.), .)
  ))

```

Replace wrong value

```{r}
fia_csv<- fia_csv %>%
  mutate(
    Corner_1.Latitude = ifelse(Polygon_ID == "P112", -1.96886, Corner_1.Latitude),
    Corner_4.Latitude = ifelse(Polygon_ID == "P104", -2.02194, Corner_4.Latitude),
     Corner_2.Longitude = ifelse(Polygon_ID == "GMT004", 29.38399, Corner_2.Longitude)
  )

```

```{r}
# Convert FIA plot data to sf object and reproject
fia_plots <- st_as_sf(fia_csv, coords = c("Central_coordinate_Longitude", "Central_coordinate_Latitude"), crs = 4326) %>%
  st_transform(crs = rwanda_crs)
```

```{r}
library(dplyr)

# Remove plots P059 and P095
fia_csv <- fia_csv %>%
  filter(!Polygon_ID %in% c("P059", "P095"))

```

# Recalculate central coordinates as the mean of four corners

```{r}

# Recalculate central coordinates as the mean of four corners
fia_csv  <- fia_csv %>%
  mutate(
    Central_coordinate_Latitude = rowMeans(select(., Corner_1.Latitude, Corner_2.Latitude, Corner_3.Latitude, Corner_4.Latitude), na.rm = TRUE),
    Central_coordinate_Longitude = rowMeans(select(., Corner_1.Longitude, Corner_2.Longitude, Corner_3.Longitude, Corner_4.Longitude), na.rm = TRUE)
  )

# Convert to sf using new central coordinates
fia_sf <- st_as_sf(fia_csv , coords = c("Central_coordinate_Longitude", "Central_coordinate_Latitude"), crs = 4326)

# Transform to Rwanda UTM CRS
fia_plots <- st_transform(fia_sf, crs = 32735)

```

```{r}

# Define typical Rwanda coordinate bounds
library(dplyr)

# Define Rwanda's latitude and longitude bounds
lat_bounds <- c(-2.9, -1.0)
lon_bounds <- c(28.8, 30.9)

# Add flag column
fia_flagged <- fia_csv  %>%
  mutate(
    coord_flag = ifelse(
      Corner_1.Latitude < lat_bounds[1] | Corner_1.Latitude > lat_bounds[2] |
      Corner_2.Latitude < lat_bounds[1] | Corner_2.Latitude > lat_bounds[2] |
      Corner_3.Latitude < lat_bounds[1] | Corner_3.Latitude > lat_bounds[2] |
      Corner_4.Latitude < lat_bounds[1] | Corner_4.Latitude > lat_bounds[2] |
      Central_coordinate_Latitude < lat_bounds[1] | Central_coordinate_Latitude > lat_bounds[2] |

      Corner_1.Longitude < lon_bounds[1] | Corner_1.Longitude > lon_bounds[2] |
      Corner_2.Longitude < lon_bounds[1] | Corner_2.Longitude > lon_bounds[2] |
      Corner_3.Longitude < lon_bounds[1] | Corner_3.Longitude > lon_bounds[2] |
      Corner_4.Longitude < lon_bounds[1] | Corner_4.Longitude > lon_bounds[2] |
      Central_coordinate_Longitude < lon_bounds[1] | Central_coordinate_Longitude > lon_bounds[2],
      "Suspicious", "OK"
    )
  )

# Save flagged dataset to CSV
write.csv(fia_flagged, "fia_flagged_coordinates2.csv", row.names = FALSE)

# Optionally print summary
table(fia_flagged$coord_flag)


```

Updates\

```{r}
library(ggplot2)
library(sf)
library(ggspatial)
library(patchwork)
library(grid)
```

\
Updates map

```{r}
# Map of Rwanda with Study Area (no FIA plots)
fig_rwanda <- ggplot() +
  geom_sf(data = rwanda, fill = "white", color = "black") +
  geom_sf(data = gmb_reserve, fill = "gray60", color = "black") +
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "tr", which_north = "true",
                         style = north_arrow_fancy_orienteering) +
  labs(
    title = "Fig. 1: Study Region within Rwanda",
    subtitle = "Shaded area shows the Gishwati-Mukura Biosphere Reserve"
  ) +
  theme_minimal()

# Print map
print(fig_rwanda)

```

### **Zoomed-in Study Area with FIA Plots**

This figure is focused for local detail and includes plot locations.

```{r}
# Zoomed-in Gishwati-Mukura Reserve Map with FIA Plots
fig_study_area <- ggplot() +
  geom_sf(data = gmb_reserve, fill = "gray60", color = "black") +
  geom_sf(data = fia_plots, color = "black", size = 2) +
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "tr", which_north = "true",
                         style = north_arrow_fancy_orienteering) +
  labs(
    title = "Fig. 2: Gishwati-Mukura Reserve (Zoomed)",
    subtitle = "Black dots represent FIA plots"
  ) +
  theme_minimal()

# Print zoomed-in study area
print(fig_study_area)

```

Deep analysis

```{r}
str(fia_csv)
```

## **Step 1: Define Site Boundaries**

## 

`gishwati.shp`

`mukura.shp`

Use them as the spatial extent for extraction.

**Step 2: Obtain Spatial Datasets**

A. **Soil and Habitat Data**

You need **vector data** (ideally from Rwandan or global sources):
**Soil**: FAO's **Harmonized World Soil Database (HWSD)** or
**SoilGrids** (global, high-resolution)

-   Download: <https://soilgrids.org>

    **Habitat/Land Cover**: **Copernicus Land Cover (100m)** or **ESA
    WorldCover (10m)**: <https://esa-worldcover.org>

    Rwanda-specific land cover: Check **Rwanda Land Management and Use
    Authority (RLMUA)** or Rwanda Spatial Data Portal

    B. **Climate Data**

Use **WorldClim v2.1** for:**Temperature**

**Precipitation**

Resolution: up to 1 km, but can be resampled.
Download: https://www.worldclim.org/data/worldclim21.html

Step 3. Process Data in R

```{r}
library(sf)
library(terra)
library(dplyr)

# Load shapefiles
gishwati <- st_read("path_to/Gishwati_core.shp")
mukura <- st_read("path_to/Mukura_core.shp")
study_area <- st_union(gishwati, mukura)

# Transform to common CRS (e.g., UTM 35S)
study_area <- st_transform(study_area, crs = 32735)

```

A. Download and Clip Soil Raster (e.g., from SoilGrids)

```{r}
# Load a raster soil layer (e.g., organic carbon or texture)
soil_raster <- rast("path_to/soil_data.tif")
soil_raster <- project(soil_raster, study_area)
soil_crop <- crop(soil_raster, vect(study_area))
soil_mask <- mask(soil_crop, vect(study_area))

```

B. **Download and Clip Habitat/Land Cover**

```{r}
landcover <- rast("path_to/landcover.tif")
landcover <- project(landcover, study_area)
landcover_crop <- crop(landcover, vect(study_area))
landcover_mask <- mask(landcover_crop, vect(study_area))

```

C. Download and Clip WorldClim Temperature & Precipitation

```{r}
# Load WorldClim Bio variables
temp <- rast("wc2.1_30s_tavg.tif")  # Mean temp
prec <- rast("wc2.1_30s_prec.tif")  # Precipitation

temp_crop <- crop(temp, vect(study_area))
temp_mask <- mask(temp_crop, vect(study_area))

prec_crop <- crop(prec, vect(study_area))
prec_mask <- mask(prec_crop, vect(study_area))

```

Step 4: Rasterize Vector Soil & Habitat to 20m Resolution

```{r}
# Create a 20x20m template raster
template <- rast(extent(study_area), resolution = 20, crs = crs(study_area))

# Rasterize a vector soil or habitat shapefile to 20m grid
soil_vector <- st_read("path_to/soil_vector.shp")
habitat_vector <- st_read("path_to/habitat_vector.shp")

soil_raster20 <- rasterize(vect(soil_vector), template, field = "soil_type")
habitat_raster20 <- rasterize(vect(habitat_vector), template, field = "habitat_type")

```

Step 5: Export Processed Rasters or Extract for Plots

```{r}
writeRaster(soil_raster20, "soil_20m.tif", overwrite = TRUE)
writeRaster(habitat_raster20, "habitat_20m.tif", overwrite = TRUE)
writeRaster(temp_mask, "temperature.tif", overwrite = TRUE)
writeRaster(prec_mask, "precipitation.tif", overwrite = TRUE)

```

Or extract data for plot coordinates:

```{r}
fia_coords <- st_as_sf(fia_csv, coords = c("Longitude", "Latitude"), crs = 4326)
fia_coords <- st_transform(fia_coords, crs = 32735)

fia_csv$soil <- terra::extract(soil_raster20, vect(fia_coords))$soil_type
fia_csv$habitat <- terra::extract(habitat_raster20, vect(fia_coords))$habitat_type
fia_csv$temp <- terra::extract(temp_mask, vect(fia_coords))[,2]  # mean temp
fia_csv$prec <- terra::extract(prec_mask, vect(fia_coords))[,2]  # total prec

```

**Updates and provide a way to get this data because will be used as
variables to corrate with invasive disstribution data**

## **Load Required Packages**

```{r}
library(terra)
library(sf)
library(exactextractr)  # For raster value extraction at points

```

##  **Import Data**

```{r}
# Park boundaries (merge Gishwati and Mukura if separate)
gishwati <- st_read("gishwati_shapefile.shp")
mukura <- st_read("mukura_shapefile.shp")
park_boundary <- st_union(gishwati, mukura)  # Combined park polygon

# WorldCover land cover raster (download ESA tile for Rwanda)
worldcover <- rast("ESA_WorldCover_2023_Rwanda.tif")

```

##  **Reclassify WorldCover Data**

Identify WorldCover codes (e.g., **`10`** = cropland, **`30`** =
herbaceous vegetation/pasture, **`50`** = urban/built-up):

```{r}
# Create masks for target classes
developed <- classify(worldcover, cbind(c(50), 1))  # Built-up areas
pasture_crop <- classify(worldcover, cbind(c(10,30), 1))  # Crops + pasture

```

## **Calculate Distance Rasters**

```{r}
# Distance to nearest developed area
dist_developed <- distance(developed, park_boundary, unit = "m")

# Distance to nearest pasture/crop area
dist_pasture_crop <- distance(pasture_crop, park_boundary, unit = "m")

```

## **Extract Distances for Invasive Species Points**

Assume invasive species data is a **`sf`** object with coordinates:

```{r}
invasive_species <- st_read("invasive_species_occurrences.shp")

# Extract distance values at occurrence points
invasive_species$dist_developed <- exact_extract(
  dist_developed, 
  invasive_species, 
  'mean'  # Use 'mean' if points overlap multiple raster cells
)

invasive_species$dist_pasture_crop <- exact_extract(
  dist_pasture_crop, 
  invasive_species, 
  'mean'
)

```

## **Export for Correlation Analysis**

```{r}
# Save as CSV for statistical software
write.csv(st_drop_geometry(invasive_species), "invasive_species_with_distances.csv")

# Optional: Save distance rasters for spatial modeling
writeRaster(dist_developed, "distance_to_developed.tif")
writeRaster(dist_pasture_crop, "distance_to_pasture_crop.tif")

```
